Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-10-10 paper/abstract.tex
somechanges
adding a bunch of paper files -- not sure why these weren't staged?

The rotational evolution of cool dwarfs is mostly unconstrained after around
2-3 billion years, and the gyrochronology relations are poorly calibrated
beyond this time due to a lack of precise ages and rotation periods for old
main-sequence stars.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, contradicting a long-held assumption that underlies many
gyrochronology relations.
We use the velocities of populations of cool dwarfs to reveal the rotational
evolution of these stars at old ages.
We find that groups of stars, selected to be the same age using a
gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
We interpret this to mean that these cooler stars are older than the hotter
ones, which implies that the period-color relation of Praesepe is not
appropriate for old stars.
Futhermore, our results indicate that the slope of the period-color relation
actually {\it inverts}: at young ages early K stars rotate more rapidly than
late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
These observations could be reproduced by mass-dependent dynamical heating,
where lower-mass stars experience a greater rate of dynamical heating than
higher-mass stars, however we find no evidence for mass-dependent heating
within K and M dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
their rotational evolution and demonstrate that kinematics have potential to
be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars lose angular momentum more quickly than
higher mass stars and therefore rotate more slowly at the same age.
However, at old ages, late G and early K dwarfs rotate {\it faster} than late
K dwarfs of the same age.
This implies that low-mass stars lose angular momentum more slowly than
higher-mass stars after a certain age or Rossby number.
Our results, based on field star rotation periods, align with recent findings
from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

% We find that groups of stars, selected to be the same age using a
% gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
% the same velocity dispersion at all temperatures; cooler stars have greater
% velocity dispersions than hotter stars.
% We interpret this to mean that these cooler stars are older than the hotter
% ones, which implies that the period-color relation of Praesepe is not
% appropriate for old stars.
% Futhermore, our results indicate that the slope of the period-color relation
% actually {\it inverts}: at young ages early K stars rotate more rapidly than
% late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
% These observations could be reproduced by mass-dependent dynamical heating,
% where lower-mass stars experience a greater rate of dynamical heating than
% higher-mass stars, however we find no evidence for mass-dependent heating
% within K and M dwarfs in the \kepler\ field.

### 2019-10-10 paper/intro.tex
adding a bunch of paper files -- not sure why these weren't staged?

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated using
their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
which is particularly useful since the ages of MS stars are difficult to
measure via isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
be caused by the deeper convective zones, and therefore stronger magnetic
dynamos (and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is often used as a
mass proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.},
and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
Not all gyrochronology models follow this form, particularly those motivated
based on stellar structure and evolution theory or with a Rossby number,
rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
vansaders2016]}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent \citep{curtis2019}.
This cluster has a flattened relationship between rotation period and color
and the G dwarfs rotate at the same rate as the K dwarfs.

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
However, although this relation accurately describes the rotation periods of F
and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
with available rotation periods), it over-predicts the rotation periods of K
dwarfs in the 1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is largely unknown.
The oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
evolution.

In this paper, gyrochronal ages of cool field dwarfs in the \kepler\ field
were calculated with the \citep{angus2019} gyrochronology model, using
de-reddened \Gaia\ \gcolor\ color and rotation periods reported by \mct\ (we
explain our data selection process in section \ref{sec:the_data}).
These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
figure \ref{fig:age_gradient}.
The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
This provides the first hint that the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, predicts the oldest stars in this sample to
be K dwarfs, it is probably either under-predicting M dwarf ages or
over-predicting K dwarf ages.
% In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
% slowly than the model predicts.
In what follows, we use a population-based stellar age indicator, velocity
dispersion, to investigate the gyrochronology relations at old ages in
the field.

>>>>

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}, however the exact nature and form of rotational
evolution is still being actively studied.
Once rotational evolution is well understood, rotation periods can be used to
precisely and accurately date stars via gyrochronology
\citep[\eg][]{kawaler1989, pinsonneault1989, barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
It is also well established that stars of the same age but different masses
have different rotation periods \citep[\eg][]{kraft1967, matt2012}, which is
thought to be caused by the variation in convection zone depth.
For example, lower mass stars with deeper convections zones are expected to
have stronger magnetic dynamos and lose angular momentum more rapidly.
However, as new rotation period measurements have become available, a more
complicated picture of stellar rotational evolution is emerging.
% However, an underlying assumption behind many empirical gyrochronology
% relations is that the relationships between rotation period and photometric
% color\footnote{As the directly observable quantity, color is often used as a
% mass proxy, and empirical gyrochronology relations are usually calibrated in
% color, rather than mass or effective temperature.},
% and rotation period and age are {\it separable}, meaning that the
% period-color relation is the same at all ages and the period-age relation is
% the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
% meibom2011, angus2015, angus2019}.
% Not all gyrochronology models follow this form, particularly those based on
% stellar structure and evolution theory or with a Rossby number, rather than
% rotation period dependence \citep[\eg][]{barnes2010, matt2012,
% vansaders2016]}.
For example, it was recently shown that old field stars rotate more rapidly
than models predicted \citep{angus2015, vansaders2016, vansaders2018,
metcalfe2019}, and a mass-dependent modification to the classical \prot\
$\propto t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} was required to
reproduce the data \citep{vansaders2016, vansaders2018}.
Even more recent analyses of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is mass and
time-dependent \citep{curtis2019 spada2019}.
This cluster has a flattened relationship between rotation period and
color: the G dwarfs rotate at the same rate as the K dwarfs.
New theoretical models that vary the rate of magnetic braking and the
redistribution of angular momentum in the interior of stars are able to
successfully reproduce the flattened period-color relation seen in old open
clusters \citep{spada2019}.
With the advent of large time-series photometry surveys, providing large
numbers of precise rotation period measurements over the last few years, we
now know that the rotational evolution of stars is a complicated process.
However, relying on open clusters (which are typically young) and
asteroseismic stars (which are typically G-type stars or more massive) to
calibrate gyrochronology limits the mass and age coverage of the calibration
sample.

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation in color and age, calibrated using the period-color
relation of Praesepe (in \gaia\ \gcolor\ color) and the period-age relation of
Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
However, although this relation accurately describes the rotation periods of F
and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
with available rotation periods), it over-predicts the rotation periods of K
dwarfs in the 1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is largely unknown.
The oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
evolution.

We calculated gyrochronal ages of cool field dwarfs in the \kepler\ field
using the \citep{angus2019} gyrochronology model, with de-reddened \Gaia\
\gcolor\ color and rotation periods reported by \mct\ (we explain our data
selection process in section \ref{sec:the_data}).
These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
figure \ref{fig:age_gradient}.
The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
This provides the first hint that the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, predicts the oldest stars in this sample to
be K dwarfs, it is probably either under-predicting M dwarf ages or
over-predicting K dwarf ages.
% In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
% slowly than the model predicts.
In what follows, we use a population-based stellar age indicator, velocity
dispersion, to investigate the gyrochronology relations at old ages in
the field.

### 2019-10-10 paper/results.tex
adding a bunch of paper files -- not sure why these weren't staged?

The constant velocity dispersion of young stars as a function of temperature

@@

In contrast however, velocity dispersion increases as a function of
temperature for old stars, meaning the \citet{angus2019} gyrochronology
relation either under-predicts the ages of old, late-K dwarfs, or
over-predicts the ages of old early-K and late-G dwarfs\footnote{Since \vb\
velocity dispersion only provides relative and not absolute ages, it is
difficult to tell whether the ages of cool stars are being under-predicted,
the ages of hot stars being are over-predicted, or both.}.
This indicates that the relationship between rotation period and photometric

@@

The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge.
The so-called `M dwarf dip' \citep{vansaders2018} is reflected in the lines of
constant velocity dispersion (and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of the upper edge of rotation period measurements is created by a
detection limit, it could indicate the rotation periods at which stars of
different temperatures become relatively inactive (and their rotation periods
therefore become undetectable).
% Once inactive, stars would have little photometric variability induced by
% star spots and their rotation periods would be difficult to measure, so may
% not feature in the \mct\ catalog.
If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
$\sim$4500 K become inactive at a similar age, and stay active longer than
stars hotter than $\sim$4500 on average.

>>>>

The flat velocity dispersion of young stars as a function of temperature

@@

If the gyrochronology relations worked at all ages and temperatures, the
bottom panel of figure \ref{fig:age_cut} would show a flat relationship
between velocity dispersion and \teff at all ages.
However, velocity dispersion {\it increases} as a function of temperature for
old stars, meaning the \citet{angus2019} gyrochronology relation either
under-predicts the ages of old, late-K dwarfs, or over-predicts the ages of
old early-K and late-G dwarfs\footnote{Since \vb\ velocity dispersion only
provides relative and not absolute ages, it is difficult to tell whether the
ages of cool stars are being under-predicted, the ages of hot stars being are
over-predicted, or both.}.
This suggests that the relationship between rotation period and photometric

@@

The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge.
The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
period catalog characterized by a dearth of slowly rotating M dwarfs between
$\sim$2750 and 4250 K, is reflected in the lines of constant velocity
dispersion (and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of the upper edge of rotation period measurements is created by a
detection limit, it could indicate the rotation periods at which stars of
different temperatures become relatively inactive (and their rotation periods
therefore become undetectable).
% Once inactive, stars would have little photometric variability induced by
% star spots and their rotation periods would be difficult to measure, so may
% not feature in the \mct\ catalog.
If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
$\sim$4500 K become inactive at a similar age, and stay active longer than
stars hotter than $\sim$4500 on average.
